---
title: "New Zeland's Best Retirement Locations"
author: "New Zeland Herald, Chris Knox (chris.knox@nzherald.co.nz)"
date: "February 2018"
output:
  html_document:
    code_folding: show
    echo: TRUE
    warning: FALSE
    message: FALSE
    theme: sandstone
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
subtitle: Preprocessing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RPostgreSQL)
drv <- dbDriver("PostgreSQL")
db <- dbConnect(drv, dbname='retirement')
```

```{r, echo=FALSE}
project_name <- "best-retirement-ta"
r_version <- "3.4.3"
```

## Intro

This is a playful analysis that ranks New Zealand's Territorial Authorities based
on which one is the best location to retire.

This is the first release of data and methods by the _New Zealand Herald's_ data journalism team.
The release is made to both enable transparency and provide resources for others interested
in data analysis in New Zealand. Please submit any comments or corrections as GitHub issues on
the projects [GitHub page](https://github.com/nzherald/best-retirement-ta)

The code used in the analysis here should work on Linux and OS X machines with the
appropriated dependencies installed. But it is unlikely to work on a Windows computer.

## Datasets

The following datasets were used:

– Average rates paid in 2016 
– Burglaries per capita in 2017
– The average annual sunshine between 1973 and 2013
– The proportion and size of the current population that is over 65
– The number of DHB staff per capita in 2017
– The cost and availability of 2 bedroom properties in 2017

### Data sources

#### Average Rates

The average annual rates in each TA were calculated from:

- [local authority rates income.xlsx](http://insights.nzherald.co.nz/data/best-retirement-area/local%20authority%20rates%20income.xlsx)
- [households by council district for 2013.xlsx](http://insights.nzherald.co.nz/data/best-retirement-area/households%20by%20council%20district%20for%202013.xlsx)

These spreadsheets where provided to the _New Zealand Herald_ by _StatsNZ_. But the data can also be
accessed directly via [https://www.stats.govt.nz/](https://www.stats.govt.nz/)

### Burglaries

The number of burlaries per capita for each Teritorial Authority were calculated from:

- [Victimisation Time and Place](https://www.police.govt.nz/about-us/publications-statistics/data-and-statistics/policedatanz/victimisation-time-and-place)
- [Subnational population estimates](http://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7502)

Unfortunately it was necessary to manually copy the number of _Unlawful Entry With Intent/Burglary, Break and Enter_ incidents for each
TA. This data is [here](http://insights.nzherald.co.nz/data/best-retirement-area/crime.csv)

The extract of the _Subnational population estimates_ used is [here](http://insights.nzherald.co.nz/data/best-retirement-area/TABLECODE7502_Data_9020e02b-f4ab-4585-8f88-31df25edc974.csv)

### Sunshine

The average annual sunshine per TA was calculated using:

- [Sunshine hours: annual average 1972-2013](https://data.mfe.govt.nz/layer/53313-sunshine-hours-annual-average-1972-2013/)
- [Geographic Boundary files](http://archive.stats.govt.nz/browse_for_stats/Maps_and_geography/Geographic-areas/digital-boundary-files.aspx)

### Population over 65

The proportion of the population over 65 was drawn from:

- [Subnational population estimates](http://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7502)

The extract of the _Subnational population estimates_ used is [here](http://insights.nzherald.co.nz/data/best-retirement-area/TABLECODE7502_Data_9020e02b-f4ab-4585-8f88-31df25edc974.csv)

### DHB Staff per capita

The DHB staff per capita for each TA were estimated from the DHB Staff per capita for each DHB. 

The data used were:

- [Population density by workforce and DHB - Data Request - 7022018NZ Herald.pdf](http://insights.nzherald.co.nz/data/best-retirement-area/Population%20density%20by%20workforce%20and%20DHB%20-%20Data%20Request%20-%207022018NZ%20Herald.pdf)
- [Subnational population estimates - TA/AU](http://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7502)
- [Subnational population estimates - DHB](http://nzdotstat.stats.govt.nz/wbos/Index.aspx?DataSetCode=TABLECODE7509)
- [Geographic Boundary files](http://archive.stats.govt.nz/browse_for_stats/Maps_and_geography/Geographic-areas/digital-boundary-files.aspx)

The DHB population estimates are [here](http://insights.nzherald.co.nz/data/best-retirement-area/TABLECODE7509_Data_2a83a026-d55e-4a9b-9fea-29dcb037dd98.csv)
and the area unit population estimates are [here](http://insights.nzherald.co.nz/data/best-retirement-area/TABLECODE7502_Data_50c3f259-7cb0-4c27-a889-93aa4577ae57.csv)

### Two bedroom property cost and availability

The estimated median value, the total number of properties, the median sale price in 2017 and the
number of properties sold in to 2017 were provided to the _New Zealand Herald_ by [CoreLogic](http://www.qv.co.nz/).

The data used is [here](http://insights.nzherald.co.nz/data/best-retirement-area/Two%20Bedroom%20Stats%2020180209.xlsx).

## Data analysis

The overall analysis was very simple. The goal was to rank each TA according to the 6 criteria
outlined above and then then calculated the average rank for each TA. In the cases were multiple
data contributed to a single criteria then each of those datasets where ranked and the average 
rank calculated.

### Data loading

All datasets are loaded into a PostgreSQL database. This process is managed via a
[makefile](https://github.com/nzherald/best-retirement-ta/blob/master/makefile).
If you checkout the
source code of this project, and have PostgreSQL installed just run
`make .o/load` - __DO NOT do this if you have a database called `retirement` already__

The loading steps just import the shape, raster and CSV files into a database. All
subsequent analysis is carried out within the database and the SQL scripts are included
here for reference.


### Average Rates, Burglary per capita, and population over 65

These calculations were all simple and can be seen here:

[__Rates__](https://github.com/nzherald/best-retirement-ta/blob/master/analysis/avg-rates.sql)

```{sql, eval=FALSE}
CREATE TABLE rates AS
SELECT d.ta, round(r.y2016 * 1e3 / d.count) as avg_rates
FROM dwellings d
JOIN total_rates r
  ON lower(d.ta) = lower(replace(r.ta, ' Council', ''))
;
```

[__Burglary__](https://github.com/nzherald/best-retirement-ta/blob/master/analysis/crime-percapita.sql)
```{sql, eval=FALSE}
CREATE OR REPLACE VIEW crime_percap AS
SELECT c.ta, round(c.burglary / value * 100000) AS burglary
FROM crime c
LEFT join ta_pop p
  ON lower(c.ta) = lower(p.area)
WHERE age = 'Total people, age'
  AND year = '2017'
  ;
```


[__Over 65__](https://github.com/nzherald/best-retirement-ta/blob/master/analysis/over-65.sql)

```{sql, eval=FALSE}

CREATE OR REPLACE VIEW over65 AS
SELECT r.area as ta, r.value as over65_pop, round(r.value / t.value * 100,1) as over65_pc
FROM ta_pop r
JOIN ta_pop t
  ON r.area = t.area
  AND r.year = t.year
JOIN rates rr ON
  lower(rr.ta) = lower(t.area) -- JUST filter out boards 
WHERE r.age = '65 years and over'
  AND t.age = 'Total people, age'
  AND r.year = '2016'
;
```


### Two bedroom properties

No actual calculation was needed for this data.

### [Sunshine](https://github.com/nzherald/best-retirement-ta/blob/master/analysis/ta-sunshine.sql)

PostGIS was used to clip the national sunshine data provided by MfE to each TA
and then extract the average value across the TA.

```{sql, eval=FALSE}
CREATE MATERIALIZED VIEW ta_sunshine AS 
WITH _s AS (
  SELECT ta2017_nam, st_summarystats(st_clip(rast, geom)) aS stats
  FROM ta2017_gv_clipped, sunshine
)
SELECT ta2017_nam, round((stats).mean) AS mean
FROM _s
WHERE ta2017_nam != 'Area Outside Territorial Authority'
;
```


### DHB staff

This is probably the weakest and most complicated portion of the analysis. It is also the
portion of the analysis that may be best replaced with some other measure.

The process was:

- Use DHB population to calculate a nominal number of DHB employees based on the employees per
  capita.
- Assume that DHB staff are spread equally across all Area Units (which they certainly are not)
  and estimate a DHB staff per area unit value.
- Where an area unit is in multiple DHBs calculate a fraction of the population that in each
  DHB using the Area Unit and DHB geographies. This fraciton is used to weight the contribution
  of each DHB to the estimated number of staff in the Area Unit
- Sum up the DHB staff in each TA from the Area Units
- Divide the estiamted DHB staff for each TA by the population of the TA.

https://github.com/nzherald/best-retirement-ta/blob/master/analysis/dhb-ta-overlap.sql

```{sql, eval=FALSE}
CREATE MATERIALIZED VIEW dhb_au_overlap AS
WITH _a AS (
SELECT 
    dhb2015_na
  , au2017 || ' ' || au2017_nam as au
  , round((st_area(st_intersection(au.geom, dhb.geom)) / (au.area_sq_km * 1e6))::numeric, 3) as overlap
FROM au2017_gv_clipped au
INNER JOIN "district-health-board-2015" dhb
ON ST_Intersects(au.geom, dhb.geom)
WHERE dhb2015_na != 'Area outside District Health Board'
)
SELECT *, rank() over (partition by au order by overlap)
FROM _a
where overlap > 0
;
```

https://github.com/nzherald/best-retirement-ta/blob/master/analysis/healthcare_professionals.sql

```{sql, eval=FALSE}
CREATE OR REPLACE VIEW healthcare_professionals_nominal AS
SELECT 
    p.dhb
  , p.all_staff * d.value / 100000 as all_staff
  , p.senior_medical * d.value / 100000 as senior_medical
  , p.junior_medical * d.value / 100000 as junior_medical
  , p.care_support * d.value / 100000 as care_support
  , p.nurses * d.value / 100000 as nurses
  , p.allied * d.value / 100000 as allied
  , d.value AS population
FROM healthcare_professionals p
LEFT JOIN dhb_pop d
  ON p.dhb = d.area
WHERE age = 'Total people, age'
AND year = '2016'
;

CREATE OR REPLACE VIEW healthcare_au AS
WITH _pop AS (
  SELECT au.*, p.value AS au_pop, d.value AS dhb_pop,
  p.value / d.value * overlap as weight
  FROM dhb_au_overlap au
  JOIN au_pop p
    ON au.au = p.area
  JOIN dhb_pop d
    ON au.dhb2015_na = d.area
  WHERE d.year = '2016'
    AND d.age = 'Total people, age'
)
SELECT h.dhb
  , a.au
  , a.au_pop
  , all_staff * weight as all_staff
  , senior_medical * weight as senior_medical
  , junior_medical * weight as junior_medical
  , care_support * weight as care_support
  , nurses * weight as nurses
  , allied * weight as allied
FROM _pop a
INNER JOIN healthcare_professionals_nominal h
  ON a.dhb2015_na = h.dhb
;

CREATE OR REPLACE VIEW healthcare_ta_nominal AS
SELECT ta2017_label
  , sum(all_staff) AS all_staff
  , sum(senior_medical) AS senior_medical
  , sum(junior_medical) AS junior_medical
  , sum(care_support) AS care_support
  , sum(nurses) AS nurses
  , sum(allied) AS allied
FROM healthcare_au a
INNER JOIN ta_au_lookup l
  ON a.au = l.au
GROUP BY ta2017_label
;

CREATE OR REPLACE VIEW healthcare_ta_percap AS
WITH _pop AS (
  SELECT area, value
  FROM ta_pop
  WHERE year = '2016' and age = 'Total people, age'
)
SELECT ta2017_label
  , round(all_staff / p.value::numeric * 100000) as all_staff
  , round(senior_medical / p.value::numeric * 100000) as senior_medical
  , round(junior_medical / p.value::numeric * 100000) as junior_medical
  , round(care_support / p.value::numeric * 100000) as care_support
  , round(nurses / p.value::numeric * 100000) as nurses
  , round(allied / p.value::numeric * 100000) as allied
FROM healthcare_ta_nominal
JOIN _pop p on lower(area) = lower(ta2017_label)
WHERE ta2017_label not in ('Chatham Islands Territory')
;
```



## Ranking

The rankings were as follows:

__Rates__

- Rank ascending based on average rates

__Burglary__

- Rank ascending based on burglaries per capita

__Sunshine__

- Rank descending based on sunshine hours

__Population__

- Rank descending based on proportion of population over 65
- Rank descending based on total size of population over 65

__Two bedroom property__

- Rank ascending based on median value
- Rank descending based on total properties
- Rank ascending based on median price in 2017
- Rank descending based on volume of sales in 2017

__DHB Staff__

- Ranks descending on total staff per capita
- Ranks descending on senior medical staff per capita
- Ranks descending on junior medical staff per capita
- Ranks descending on care and support staff per capita
- Ranks descending on nursing staff per capita
- Ranks descending on allied and scientific staff per capita


```{sql, eval=FALSE}
CREATE or replace view health_rank AS
WITH _r AS (
  SELECT ta2017_label
   , rank() over (order by all_staff desc) as all_staff
   , rank() over (order by senior_medical desc) as senior_medical
   , rank() over (order by junior_medical desc) as junior_medical
   , rank() over (order by care_support desc) as care_support
   , rank() over (order by nurses desc) as nurses
   , rank() over (order by allied desc) as allied
  FROM healthcare_ta_percap
)
select ta2017_label as ta, rank() over (order by (all_staff + senior_medical + junior_medical + care_support + nurses + allied)::numeric / 6)
FROM _r;

create or replace view over65_rank as
with _inner as (
  select ta, rank() over (order by over65_pop desc) as pop, rank() over (order by over65_pc desc) as pc
  from over65
)
select ta, rank() over (order by pop + pc)
from _inner;

create or replace view property_rank as
with _inner as (
  select name as ta
  , rank() over (order by total_sales_in_2017 desc) as sales
  , rank() over (order by median_sales_price) as price
  , rank() over (order by total_properties desc) as total
  , rank() over (order by median_value) as value
  FROM two_bedroom
)
SELECT *, rank() over (order by (sales + price + total + value)::numeric / 4)
FROM _inner;

create or replace view rank as
with _inner AS (
select r.ta
  , rank() over (order by avg_rates) as rates
  , rank() over (order by mean desc) as sunshine
  , rank() over (order by burglary) as burglary
  , o.rank as over65
  , h.rank as health
  , p.rank as property
from rates r
left join ta_sunshine s
  on r.ta = s.ta2017_nam
left join over65_rank o
  on lower(r.ta) = lower(o.ta)
left join health_rank h
  on r.ta = h.ta
left join property_rank p
  on r.ta = p.ta
left join crime_percap c
  ON lower(r.ta) = lower(c.ta)
)
select *, rank() over (order by (rates + sunshine + over65 + health + property + burglary)::numeric / 6) as overall
from _inner
;
```



